### 2021-06-03 papers/thesis/full/ch4-spherical-caps/spherical-caps.tex
Further edits. New spy plots to make uniform across thesis. New references and fixing of capitals etc.

The spherical cap can be thought of as higher dimensional version of the circle arc (a one-dimensional \enquote{surface} in two-dimensional space). One can define OPs on the circle in terms of Fourier series, which we write here as orthogonal polynomials in $x$ and $y$:
\begin{definition}\label{def:sc:Ydefinition}
	Define the unit circle $\omega := \bsset{\vec{x} = (x,y) \in \R^2}{x^2 + y^2 = 1}$, and define the parameter $\theta$ for each $(x,y) \in \omega$ by $x = \cos\theta$, $y = \sin\theta$. Define the polynomials $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ on $(x,y) \in \omega$ by
\bseqn{
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) &:= \ch_{0} =: \ch_{0,0}(\theta) \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) &:= T_k(x) = \cos k \theta =: \ch_{k,0}(\theta) \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) &:= y \: U_{k-1}(x) = \sin k \theta =: \ch_{k,1}(\theta)
}
for $k = 1,2,3,\dots$ where $\ch_{0} := {\sqrt{2} \over 2}$ and $T_k$, $U_{k-1}$ are the standard Chebyshev polynomials on the interval $[-1,1]$. The $\{Y_{k,i}\}$ are orthonormal with respect to the inner product
\bseqn{
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta.
}
Note that $\ch_0$ is defined so as to ensure orthonormality. 
\end{definition}

@@

\bstodoinline{Expand on this. Read the Olver Xu "OPs on quadratic curves paper." where the arc might apply??? Either way, cite this paper in intro (CH1).}

\section{Orthogonal polynomials on algebraic curves}

This idea of using 1D OPs on a shortened interval to create the modified OPs for a circle arc has be generalised to higher dimensional domains, where a way of constructing orthogonal polynomials on algebraic curves has been presented \cite{olver2020orthogonal}. \bstodoinline{Expand on this. Summarise the "on" surfaces approach. Use references in the paper.}

@@

Using the approach presented for constructing OPs on algebraic curves, we can construct the 3D orthogonal polynomials on the spherical cap or band $\Omega$ from 1D orthogonal polynomials on the interval $[\alpha,\beta]$, and from Fourier series written as orthogonal polynomials in $x$ and $y$. Recall that the spherical cap is defined by 
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{\alpha < z < \beta, \: x^2 + y^2 + z^2 = 1}.
}
\begin{proposition}[\cite{olver2020orthogonal}]\label{prop:sc:construction}
	Let $w : (\alpha,\beta) \to \R$ be a weight function. For $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$ given by
\bseqn{
	\scopnki(x,y,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr({x \over \rho(z)}, {y \over \rho(z)})
}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\bseqn{
	&\ip< p, \: q > := \int_\Omega p(x,y,z) \: q(x,y,z) \: w(z) \: \D A \\
	&= \int_0^{\cos^{-1}(\alpha)} \int_0^{2\pi} \Big\{ p\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \: q\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \\
	&\quadeight \quad \cdot w(\cosphi) \: \sinphi \Big\} \: \D \theta \: \D \varphi \\
	&= \int_\alpha^1 \int_0^{2\pi} p\big(\rho(z) \costheta, \rho(z) \sintheta, z\big) \: q\big(\rho(z) \costheta, \rho(z) \sintheta, z\big) \: w(z) \: \D \theta \: \D z
}
on $\Omega$, where $\D A = \sinphi \: \D \theta \: \D \varphi$ is the uniform spherical measure on $\Omega$. 
\end{proposition}

@@

\begin{figure}[tp]
	\centering % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-laplacian}}
		\centering
		\caption{The Laplace-Beltrami operator $\Delta^{(1)}_W$}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-rho2laplacian}}
		\centering
		\caption{The $\rho^2$-factored Laplace-Beltrami operator $D_\varphi^{(0)} \: W_\varphi^{(1)} + T^{(0)\to(1)} \: T_W^{(1)\to(0)} \: (D_\theta)^2$}
	\end{subfigure}\hfil % <-- added

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-helmholtz}}
		\centering
		\caption{A variable coefficient Helmholtz operator}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-biharmonic}}
		\centering
		\caption{The Biharmonic operator $\calB_W^{(2)}$} 
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of (differential) operator matrices, showing their sparsity. For (c), the weighted variable coefficient Helmholtz operator is $\Delta^{(1)}_W + k^2 \: T^{(0)\to(1)} \: V({J_x^{(0)}}^\top, {J_y^{(0)}}^\top, {J_z^{(0)}}^\top) \: T_W^{(1)\to(0)}$ for $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 200$.}
	\label{fig:sc:sparsity}
\end{figure}

@@

Our last example is the biharmonic equation: find $u(x,y,z)$ given a function $f(x,y,z)$ such that:
\bseqnnumber{
	\begin{cases}
    		\DeltaS^2 u(x,y,z) = f(x,y,z) &\quad \text{in } \Omega \\
		u(x,y,z) = 0, \quad \frac{\partial u}{\partial n}(x,y,z) = \grad_S \: u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 &\quad \text{on } \partial \Omega
	\end{cases}
	\label{eqn:sc:biharmonic}
}
where $\DeltaS^2$ is the Biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. For clarity, we reiterate that the unit normal vector in this sense is simply $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$ (see \bsrefsection{Section:sc:furtherdiffoperators}). In \bsreffig{fig:sc:biharmonic} we see the solution to the Biharmonic \bsrefeqn{eqn:sc:biharmonic} in the spherical cap $\Omega$. In \bsreffig{fig:sc:biharmonic} we also show the norms of each block of calculated coefficients of the approximation for four more complex right-hand sides of the biharmonic equation with $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve exponential convergence for these more complex functions.

>>>>

The spherical cap can be thought of as a higher dimensional version of the circle arc (a one-dimensional \enquote{surface} in two-dimensional space). Recently, a procedure for defining OPs in two variables on quadratic curves, such as ellipses and parabolas, has been presented \cite{olver2021orthogonal}, where the unit circle is one of the five classes of curve that can be generalised to give any curve in the plane $\R^2$. One can define OPs on the circle in terms of Fourier series, which we write here as orthogonal polynomials in $x$ and $y$:
\begin{definition}[\cite{olver2021orthogonal}]\label{def:sc:Ydefinition}
	Define the unit circle $\omega := \bsset{\vec{x} = (x,y) \in \R^2}{x^2 + y^2 = 1}$, and define the parameter $\theta$ for each $(x,y) \in \omega$ by $x = \cos\theta$, $y = \sin\theta$. Define the polynomials $\{\chki\}$ for $k = 0, 1, \dots$, $i = 0, 1$ on $(x,y) \in \omega$ by
\bseqn{
	\ch_{0,0}(\xvec) \equiv \ch_{0,0}(x, y) &:= \ch_{0} =: \ch_{0,0}(\theta) \\
	\ch_{k,0}(\xvec) \equiv \ch_{k,0}(x, y) &:= T_k(x) = \cos k \theta =: \ch_{k,0}(\theta) \\
	\ch_{k,1}(\xvec) \equiv \ch_{k,1}(x, y) &:= y \: U_{k-1}(x) = \sin k \theta =: \ch_{k,1}(\theta)
}
for $k = 1,2,3,\dots$ where $\ch_{0} := {\sqrt{2} \over 2}$ and $T_k$, $U_{k-1}$ are the standard Chebyshev polynomials on the interval $[-1,1]$. The $\{Y_{k,i}\}$ are orthonormal with respect to the inner product
\bseqn{
	\ip< p, \: q >_{\ch} &:= \frac{1}{\pi} \: \int_0^{2\pi} p(\xvec(\theta)) \: q(\xvec(\theta)) \: \D \theta.
}
Note that $\ch_0$ is defined so as to ensure orthonormality. 
\end{definition}

@@

@@

A continuation to the procedure for defining OPs on quadratic curves (see \cite{olver2021orthogonal}) has also been presented \cite{olver2020orthogonal}, where OPs are defined on quadratic surfaces of revolution in higher dimensions (specifically, surfaces in $\R^{d+1} for d \ge 2$). The authors describe their technique as modelled on the structure of the unit sphere, by approaching the construction as a generalisation of spherical harmonics.

Using this approach, we can construct the 3D orthogonal polynomials on the spherical cap or band $\Omega$ from 1D orthogonal polynomials on the interval $[\alpha,\beta]$, and from Fourier series written as orthogonal polynomials in $x$ and $y$. Recall that the spherical cap is defined by
\bseqn{
	\Omega := \bsset{(x,y,z) \in \R^3}{\alpha < z < \beta, \: x^2 + y^2 + z^2 = 1}.
}
\begin{proposition}[\cite{olver2020orthogonal}]\label{prop:sc:construction}
	Let $w : (\alpha,\beta) \to \R$ be a weight function. For $n = 0,1,2,\dots, $ let $\{r_{n,k}\}$ be polynomials orthogonal with respect to the weight $\rho(x)^{2k} w(x)$ where $0 \le k \le n$. Then the 3D polynomials defined on $\Omega$ given by
\bseqn{
	\scopnki(x,y,z) := r_{n-k,k}(z) \: \rho(z)^k \: \ch_{k,i}\fpr({x \over \rho(z)}, {y \over \rho(z)})
}
for $i \in {0,1}, \: 0 \le k \le n, \: n = 0,1,2,\dots$ are orthogonal polynomials with respect to the inner product
\bseqn{
	&\ip< p, \: q > := \int_\Omega p(x,y,z) \: q(x,y,z) \: w(z) \: \D A \\
	&= \int_0^{\cos^{-1}(\alpha)} \int_0^{2\pi} \Big\{ p\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \: q\big(\sinphi \costheta, \sinphi \sintheta, \cosphi \big) \\
	&\quadeight \quad \cdot w(\cosphi) \: \sinphi \Big\} \: \D \theta \: \D \varphi \\
	&= \int_\alpha^1 \int_0^{2\pi} p\big(\rho(z) \costheta, \rho(z) \sintheta, z\big) \: q\big(\rho(z) \costheta, \rho(z) \sintheta, z\big) \: w(z) \: \D \theta \: \D z
}
on $\Omega$, where $\D A = \sinphi \: \D \theta \: \D \varphi$ is the uniform spherical measure on $\Omega$. 
\end{proposition}

@@

\begin{figure}[tp]
	\centering % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-laplacian}}
		\centering
		\caption{}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-rho2laplacian}}
		\centering
		\caption{}
	\end{subfigure}\hfil % <-- added

	\medskip
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-helmholtz}}
		\centering
		\caption{}
	\end{subfigure}\hfil % <-- added
	\begin{subfigure}{0.5\textwidth}
		\centerline{\includegraphics[scale=0.5]{sparsity-of-biharmonic}}
		\centering
		\caption{} 
	\end{subfigure}\hfil % <-- added
	\caption{\enquote{Spy} plots of (differential) operator matrices, showing their sparsity, plotted for $N = 20$. (a) The Laplace-Beltrami operator $\Delta^{(1)}_W$. (b) The $\rho^2$-factored Laplace-Beltrami operator $D_\varphi^{(0)} \: W_\varphi^{(1)} + T^{(0)\to(1)} \: T_W^{(1)\to(0)} \: (D_\theta)^2$. (c) The variable coefficient Helmholtz operator $\Delta^{(1)}_W + k^2 \: T^{(0)\to(1)} \: V({J_x^{(0)}}^\top, {J_y^{(0)}}^\top, {J_z^{(0)}}^\top) \: T_W^{(1)\to(0)}$ for $v(x,y,z) = 1 - (3(x-x_0)^2 + 5(y-y_0)^2 + 2(z-z_0)^2)$ where $(x_0, z_0) := (0.7, 0.2)$, $y_0 := \sqrt{1 - x_0^2 - z_0^2}$ and $k = 200$. (d) The Biharmonic operator $\calB_W^{(2)}$.}
	\label{fig:sc:sparsity}
\end{figure}

@@

Our last example is the biharmonic equation: find $u(x,y,z)$ given a function $f(x,y,z)$ such that:
\bseqnnumber{
	\begin{cases}
    		\DeltaS^2 u(x,y,z) = f(x,y,z) &\quad \text{in } \Omega \\
		u(x,y,z) = 0, \quad \frac{\partial u}{\partial n}(x,y,z) = \grad_S \: u(x,y,z) \cdot \unitvec{n}(x,y,z) = 0 &\quad \text{on } \partial \Omega
	\end{cases}
	\label{eqn:sc:biharmonic}
}
where $\DeltaS^2$ is the Biharmonic operator, noting the imposition of zero Dirichlet and Neumann boundary conditions on $u$. For clarity, we reiterate that the unit normal vector in this sense is simply $\unitvec{n}(x,y,z) = \unitvec{n}(\vec x) := {\vec x \over \norm{\vec x}} = \vec x$ (see \bsrefsection{section:sc:furtherdiffoperators}). In \bsreffig{fig:sc:biharmonic} we see the solution to the Biharmonic \bsrefeqn{eqn:sc:biharmonic} in the spherical cap $\Omega$. In \bsreffig{fig:sc:biharmonic} we also show the norms of each block of calculated coefficients of the approximation for four more complex right-hand sides of the biharmonic equation with $N = 200$, that is, $(N+1)^2 = 40,401$ unknowns. Once again, the rate of decay in the coefficients is a proxy for the rate of convergence of the computed solution, and we see that we achieve exponential convergence for these more complex functions.
